---
title: "Estimating Parameters and Fitting Distributions"
subtitle: "Dormann (2020), *Environmental Data Analysis* — Chapter 3 (3.3, 3.5) and Chapter 4 (4.2–4.4)"
author: "Emiel van Loon"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    fig-width: 9
    fig-height: 4.5
    fig-align: center
    code-fold: true
    code-summary: "R code"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(MASS)
knitr::opts_chunk$set(dev = "png", dpi = 150)
set.seed(42)

# the same shared look as in presentation 1a / 2a
pal  <- c(bar = "grey80", accent = "#1f78b4", accent2 = "#e31a1c",
          accent3 = "#33a02c", dark = "grey25")
parset <- function(...) par(mar = c(4.2, 4.2, 2.2, 1), bty = "n",
                           cex.lab = 1.1, cex.axis = 1.0, ...)
cols <- pal[c("dark", "accent", "accent2")]

lang <- nlschools$lang
iq   <- nlschools$IQ
yld  <- c(immer$Y1, immer$Y2)
wait <- geyser$waiting
```

```{=html}
<style>
.reveal .fignote { font-size: 0.60em; line-height: 1.35; }
.reveal .fignote code { font-size: 0.95em; }
</style>
```

## Where we are going {.smaller}

Last time we **chose** parameter values and looked at the resulting curve.
Now we turn the question around: given data, **which parameter values fit best?**

At the end of this chapter you should...

- know the difference between an **estimator** and an **estimate**
- understand **likelihood** as the density function read backwards
- know why we always work with the **log**-likelihood
- be able to fit a distribution in R with **`fitdistr()`** and read its output
- compare candidate distributions with **log-likelihood, AIC and BIC**
- compare a fit with the data graphically via the **ECDF**, and formally with
  the **Kolmogorov-Smirnov** test
- know why **normality tests** should not be trusted blindly

::: {.callout-note}
Illustrations use `nlschools`, `immer` and `geyser` from the `MASS` package —
which also provides `fitdistr()`.
:::

## The question, turned around

```{r turnaround}
#| fig-height: 4.0
parset(mfrow = c(1, 2))
hist(iq, breaks = 25, freq = FALSE, col = pal["bar"], border = "white",
     main = "last week: parameters given", xlab = "verbal IQ", ylab = "density")
curve(dnorm(x, 10, 2), add = TRUE, lwd = 3, col = pal["accent"])
curve(dnorm(x, 12, 4), add = TRUE, lwd = 3, col = pal["accent3"])
hist(iq, breaks = 25, freq = FALSE, col = pal["bar"], border = "white",
     main = "this week: parameters estimated", xlab = "verbal IQ",
     ylab = "density")
curve(dnorm(x, mean(iq), sd(iq)), add = TRUE, lwd = 3, col = pal["accent2"])
```

::: {.fignote}
Left: two normal curves we simply *picked*; neither is convincing.
Right: the one normal curve that makes these 2287 pupils **as likely as
possible**. Finding it is called *fitting a distribution to the data*, and it
needs two ingredients: a measure of fit, and a way to maximise it.
:::

## Estimator, estimate, and what makes one good {.smaller}

An **estimator** is a *rule* for calculating a quantity from data.
An **estimate** is the *number* that rule produces for one particular sample.

> the sample mean $\bar{x}$ is an estimator; 40.93 is the estimate

::: {.fragment}
Two things we want from an estimator:

- **consistent** — the more data, the closer the estimate comes to the truth
- **unbiased** — it is not systematically too high or too low
:::

::: {.fragment}
::: {.callout-important}
Careful with symbols: $\bar{x}$ and *s* describe a **sample** (any sample, any
distribution). $\mu$ and $\sigma$ are **parameters of a normal distribution**.
Writing $\mu$ already commits you to normality.
:::
:::

## Likelihood: the density function, read backwards {.smaller}

The density function answers: *given the parameters, how probable are these
data?*

The **likelihood** answers the mirror question: *given these data, how plausible
are those parameters?*

::: {.fragment}
The two are numerically the same quantity — only the reading direction differs.
The `d...()` functions therefore do double duty: `dnorm()`, `dbinom()`,
`dpois()` are our likelihood machines.
:::

::: {.fragment}
::: {.callout-tip}
Crucially: in a **distribution** plot the data vary along the *x*-axis and the
parameters are fixed. In a **likelihood** plot the data are fixed and the
**parameter** varies along the *x*-axis.
:::
:::

## Dormann's coin: 60 heads in 100 tosses

```{r coin-pmf}
#| fig-height: 4.0
parset()
kk <- 20:80
plot(kk, dbinom(kk, 100, 0.5), type = "l", lwd = 3, col = pal["dark"],
     xlab = "number of heads", ylab = "probability", ylim = c(0, 0.09))
lines(kk, dbinom(kk, 100, 0.6), lwd = 3, col = pal["accent"])
abline(v = 60, lty = 2, col = pal["accent2"], lwd = 2)
points(60, dbinom(60, 100, 0.5), pch = 19, cex = 1.6, col = pal["dark"])
points(60, dbinom(60, 100, 0.6), pch = 19, cex = 1.6, col = pal["accent"])
legend("topleft", bty = "n", lwd = 3, col = pal[c("dark", "accent")],
       legend = c("Binom(100, 0.5): P(60) = 0.011",
                  "Binom(100, 0.6): P(60) = 0.081"))
```

::: {.fignote}
Two candidate values of *p*, and the one observation we actually made (dashed
line). Under *p* = 0.6 our result is about **eight times** more probable than
under *p* = 0.5 — so *p* = 0.6 is the more plausible parameter.
:::

## The likelihood function: fix the data, vary the parameter

```{r coin-lik}
#| fig-height: 4.2
parset(mfrow = c(1, 2))
ps <- seq(0.001, 0.999, length.out = 1000)
L  <- dbinom(60, 100, ps)
plot(ps, L, type = "l", lwd = 3, col = pal["accent"], xlab = "p",
     ylab = "likelihood", main = "likelihood")
abline(v = ps[which.max(L)], lty = 2, col = pal["accent2"], lwd = 2)
plot(ps, log(L), type = "l", lwd = 3, col = pal["accent"], xlab = "p",
     ylab = "log-likelihood", ylim = c(-40, 0), main = "log-likelihood")
abline(v = ps[which.max(log(L))], lty = 2, col = pal["accent2"], lwd = 2)
```

::: {.fignote}
Same `dbinom()` call, completely different picture: now *p* runs along the
*x*-axis. The peak sits at **0.6** — the *maximum likelihood estimate*, and
exactly what the analytical formula $\hat{p} = \bar{x}/n$ gives. Taking the
logarithm squashes the *y*-axis but **cannot move the maximum**, because the
logarithm is strictly increasing.
:::

## From one data point to a whole data set

The data points are assumed **independent**, so the probability of observing
*all* of them is the **product** of their individual densities.

::: {.fragment}
That product is the likelihood of the data set. And it is a disaster
numerically: multiply 2287 numbers smaller than 1 and you get...

```{r underflow}
#| echo: true
#| code-fold: false
prod(dnorm(nlschools$lang, mean = 40.93, sd = 9.00))
```
:::

::: {.fragment}
Zero. Not because the data are impossible, but because the computer ran out of
decimal places. Hence the old trick: take the **logarithm**, and the product
becomes a **sum**.
:::

## Three ways to the same log-likelihood

```{r loglik-ways}
#| echo: true
#| code-fold: false
x <- nlschools$lang
fit <- fitdistr(x, "normal")
sum(log(dnorm(x, fit$estimate[1], fit$estimate[2])))   # by hand
sum(dnorm(x, fit$estimate[1], fit$estimate[2], log = TRUE))  # safer
logLik(fit)                                            # from the fit
```

::: {.fignote}
`log = TRUE` is not just shorter: it computes the logarithm *inside* the density
function, so no tiny intermediate number ever has to be stored. `logLik()`
also reports **df = 2**, the number of estimated parameters — remember that, we
need it for AIC.
:::

## Climbing the log-likelihood surface

```{r surface}
#| fig-height: 4.4
parset(mfrow = c(1, 2))
mus <- seq(39.5, 42.5, length.out = 120)
ll1 <- sapply(mus, function(m) sum(dnorm(lang, m, sd(lang), log = TRUE)))
plot(mus, ll1, type = "l", lwd = 3, col = pal["accent"],
     xlab = expression(mu), ylab = "log-likelihood",
     main = "vary the mean only")
abline(v = mus[which.max(ll1)], lty = 2, lwd = 2, col = pal["accent2"])

sds <- seq(8.0, 10.2, length.out = 80)
mus2 <- seq(40.0, 41.9, length.out = 80)
Z <- outer(mus2, sds, Vectorize(function(m, s)
  sum(dnorm(lang, m, s, log = TRUE))))
contour(mus2, sds, Z, nlevels = 25, drawlabels = FALSE, col = "grey70",
        xlab = expression(mu), ylab = expression(sigma),
        main = "vary both parameters")
points(mean(lang), sd(lang), pch = 19, cex = 1.8, col = pal["accent2"])
```

::: {.fignote}
`nlschools$lang`, *n* = 2287. Left: a smooth hill with one clear top. Right: with
two parameters the likelihood is a **surface**, and the maximum likelihood
estimate is its summit (red). Fitting a distribution is climbing this hill.
:::

## Sometimes the summit has a formula

For a few distributions the top of the hill can be found with pen and paper:

| distribution | maximum likelihood estimator |
|---|---|
| normal | $\hat{\mu}$ = the sample mean, $\hat{\sigma}$ = the sample sd |
| Poisson | $\hat{\lambda}$ = the sample mean |
| binomial (*n* known) | $\hat{p}$ = mean number of successes / *n* |

```{r analytic}
#| echo: true
#| code-fold: false
c(mean = mean(lang), sd = sd(lang))
fitdistr(lang, "normal")$estimate
```

::: {.fignote}
So the familiar sample statistics from Chapter 1 are **not arbitrary** — for the
normal distribution they *are* the maximum likelihood estimators. (The tiny
difference in the sd is the *n* versus *n* − 1 divisor.)
:::

## When there is no formula — and why we bother {.smaller}

For most distributions (the negative binomial, for instance) no closed
expression exists, so R **searches** for the summit numerically — trial and
error, guided by a fast optimisation algorithm. That is all `fitdistr()` does.

::: {.fragment}
Why maximum likelihood is the default method:

1. **asymptotically unbiased** — with more data the estimate converges on the truth
2. **consistent** — and it converges ever more precisely
3. **efficient** — among unbiased estimators it has the lowest variance
4. **transformation-friendly** — the MLE of $\sigma$ is the square root of the MLE of $\sigma^2$
:::

::: {.fragment}
The main alternative is the **method of moments** (match sample mean, variance,
... to the distribution's). Easier to compute, usually less efficient.
:::

## Consistency, seen

```{r consistency}
#| fig-height: 4.2
parset()
ns <- c(5, 10, 20, 50, 100, 250, 500, 1000, 2000)
est <- sapply(ns, function(n) replicate(300, mean(sample(lang, n))))
plot(NA, xlim = range(ns), ylim = c(33, 49), log = "x", xlab = "sample size n",
     ylab = "estimate of the mean")
for (i in seq_along(ns))
  points(rep(ns[i], 300), est[, i], pch = 19, cex = 0.35,
         col = adjustcolor(pal["accent"], 0.25))
abline(h = mean(lang), lwd = 3, col = pal["accent2"])
lines(ns, apply(est, 2, quantile, 0.025), lwd = 2, lty = 2, col = pal["dark"])
lines(ns, apply(est, 2, quantile, 0.975), lwd = 2, lty = 2, col = pal["dark"])
```

::: {.fignote}
300 samples of each size drawn from the 2287 pupils; the red line is the truth.
The cloud is **centred** on it at every *n* (unbiased) and **narrows** as *n*
grows (consistent). With *n* = 5 an estimate can easily be 6 points off.
:::

## `fitdistr()`: the estimates come with error bars

```{r fitdistr-out}
#| echo: true
#| code-fold: false
fitdistr(c(immer$Y1, immer$Y2), "normal")   # n = 60 barley plots
fitdistr(nlschools$lang, "normal")          # n = 2287 pupils
```

::: {.fignote}
The second line of each block is the **standard error** of the estimate.
The barley mean is known to about ±3.5, the pupils' mean to about ±0.19: with 38
times more data the estimate is roughly 6 times more precise (√38 ≈ 6). Always
report the estimate *with* its uncertainty.
:::

## Which distribution? First think about the process {.smaller}

Often the process tells you: counts of independent events → Poisson; *k*
successes out of *n* → binomial; many small additive contributions → normal.

::: {.fragment}
It is the awkward cases that need judgement:

- **percentage cover** of a plant: binomial? beta? and what if it was recorded on
  the coarse Braun-Blanquet scale and back-converted?
- a **ratio** of two varying counts: the distribution of the quotient is hard
  to predict
:::

::: {.fragment}
::: {.callout-warning}
The reflex use of the normal distribution is the commonest mistake. When there
is no clear candidate, fit several and let the data choose — that is what the
rest of this talk is about.
:::
:::

## Transforming the data is not a free lunch

```{r transform}
#| fig-height: 4.0
parset(mfrow = c(1, 2))
hist(yld, breaks = 15, freq = FALSE, col = pal["bar"], border = "white",
     main = "immer yields", xlab = "yield", ylab = "density")
curve(dnorm(x, mean(yld), sd(yld)), add = TRUE, lwd = 3, col = pal["accent2"])
hist(log(yld), breaks = 15, freq = FALSE, col = pal["bar"], border = "white",
     main = "log(yields)", xlab = "log yield", ylab = "density")
curve(dnorm(x, mean(log(yld)), sd(log(yld))), add = TRUE, lwd = 3,
      col = pal["accent2"])
```

::: {.fignote}
Taking logs turns a log-normal problem into a normal one — but it also
**changes the variance**, and the result must satisfy the assumptions of the
*new* distribution. The square root of Poisson data often looks normal for
exactly this reason (it stabilises the variance). Useful, but never automatic.
:::

## Comparing fits: the log-likelihood

```{r compare-fit}
#| fig-height: 4.2
parset()
fits <- list(normal    = fitdistr(yld, "normal"),
             lognormal = fitdistr(yld, "lognormal"),
             gamma     = fitdistr(yld, "gamma"),
             weibull   = fitdistr(yld, "weibull"))
hist(yld, breaks = 15, freq = FALSE, col = pal["bar"], border = "white",
     main = "", xlab = "barley yield (immer, n = 60)", ylab = "density")
cc <- c(pal["dark"], pal["accent"], pal["accent2"], pal["accent3"])
curve(dnorm(x, fits$normal$estimate[1], fits$normal$estimate[2]),
      add = TRUE, lwd = 3, col = cc[1])
curve(dlnorm(x, fits$lognormal$estimate[1], fits$lognormal$estimate[2]),
      add = TRUE, lwd = 3, col = cc[2])
curve(dgamma(x, fits$gamma$estimate[1], fits$gamma$estimate[2]),
      add = TRUE, lwd = 3, col = cc[3])
curve(dweibull(x, fits$weibull$estimate[1], fits$weibull$estimate[2]),
      add = TRUE, lwd = 3, col = cc[4])
legend("topright", bty = "n", lwd = 3, col = cc, legend = names(fits))
```

::: {.fignote}
Four distributions fitted by maximum likelihood to the same data. The one with
the **highest log-likelihood** made these 60 yields most probable — so higher
is better. But is that enough?
:::

## AIC and BIC: paying for extra parameters {.smaller}

A distribution with more parameters is more flexible and will *always* be able
to reach a higher log-likelihood. Comparing $\ell$ alone therefore rewards
complexity for its own sake.

| criterion | penalty per parameter | note |
|---|---|---|
| log-likelihood $\ell$ | none | **higher is better** |
| AIC | 2 | **lower is better** |
| AICc | slightly more than 2 | AIC corrected for small *n*; use it by default |
| BIC | log(*n*) | penalises harder as the sample grows |

::: {.fragment}
All three start from the same $\ell$ and subtract a price for each parameter you
were allowed to tune. With *n* = 60 the BIC penalty is log(60) ≈ 4.1 per
parameter — twice as strict as AIC.
:::

## The comparison table

```{r compare-table}
tab <- t(sapply(fits, function(f)
  c(logLik = as.numeric(logLik(f)), AIC = AIC(f), BIC = BIC(f))))
knitr::kable(round(tab[order(tab[, "AIC"]), ], 1))
```

::: {.fignote}
`immer` yields, *n* = 60. The log-normal wins, the gamma is a close second, and
the normal and Weibull trail. Here all four candidates have **two** parameters,
so AIC and BIC merely reorder what $\ell$ already said — the three criteria
agree. They need not always.
:::

## When the number of parameters differs

```{r compare-np}
fn <- fitdistr(lang, "normal")    # 2 parameters
fp <- fitdistr(lang, "Poisson")   # 1 parameter
tab2 <- rbind(
  `normal (2 par.)`  = c(logLik = as.numeric(logLik(fn)), AIC = AIC(fn), BIC = BIC(fn)),
  `Poisson (1 par.)` = c(as.numeric(logLik(fp)), AIC(fp), BIC(fp)))
knitr::kable(round(tab2, 1))
```

::: {.fignote}
`nlschools$lang` are whole numbers, so a Poisson is conceivable — but for a
Poisson the variance must equal the mean, and here the mean is 40.9 while the
variance is only 81... far from a match once you also demand the right shape.
The normal wins on every criterion, and comfortably enough that the one saved
parameter cannot rescue the Poisson. Exactly Dormann's ash-tree comparison
(Table 3.1).
:::

## The ECDF: the data as a cumulative curve

```{r ecdf-intro}
#| fig-height: 4.2
parset(mfrow = c(1, 2))
small <- sort(sample(yld, 12))
plot(ecdf(small), verticals = TRUE, pch = 19, col.01line = NA,
     main = "12 observations", xlab = "yield", ylab = "cumulative proportion",
     col = pal["accent"], lwd = 2)
plot(ecdf(lang), verticals = TRUE, pch = "", col.01line = NA,
     main = "2287 observations", xlab = "language score",
     ylab = "cumulative proportion", col = pal["accent"], lwd = 2)
```

::: {.fignote}
Sort the values and step up by 1/*n* at each one: that is the **empirical
cumulative distribution function**. No bin width to choose, no smoothing — every
data point is in there. With few data it is a coarse staircase; with many it
becomes a smooth curve that we can lay a theoretical cdf on top of.
:::

## Comparing an ECDF with a fitted distribution

```{r ecdf-compare}
#| fig-height: 4.2
parset(mfrow = c(1, 2))
for (nm in c("normal", "lognormal")) {
  e <- fits[[nm]]$estimate
  plot(ecdf(yld), verticals = TRUE, pch = "", col.01line = NA, lwd = 2,
       main = nm, xlab = "barley yield", ylab = "cumulative probability")
  if (nm == "normal") curve(pnorm(x, e[1], e[2]), add = TRUE, lwd = 3,
                            col = pal["accent2"])
  else                curve(plnorm(x, e[1], e[2]), add = TRUE, lwd = 3,
                            col = pal["accent2"])
  lines(ecdf(yld), verticals = TRUE, pch = "", col.01line = NA, lwd = 2)
}
```

::: {.fignote}
Black: the data. Red: the fitted cumulative distribution. The normal curve rises
too early in the left half and lags in the right tail; the log-normal tracks the
staircase more closely. This graphical comparison is often **more revealing
than any test statistic**.
:::

## The Kolmogorov-Smirnov test measures the biggest gap

```{r ks-d}
#| fig-height: 4.2
parset()
e <- fits$normal$estimate
xs <- sort(yld); n <- length(xs)
Fe <- ppoints(n, a = 0); Ft <- pnorm(xs, e[1], e[2])
gap <- abs((1:n)/n - Ft); j <- which.max(gap)
plot(ecdf(yld), verticals = TRUE, pch = "", col.01line = NA, lwd = 2,
     main = "", xlab = "barley yield", ylab = "cumulative probability")
curve(pnorm(x, e[1], e[2]), add = TRUE, lwd = 3, col = pal["accent2"])
segments(xs[j], (j)/n, xs[j], Ft[j], lwd = 4, col = pal["accent3"])
text(xs[j], (j/n + Ft[j])/2, sprintf("  D = %.3f", max(gap)), pos = 4,
     cex = 1.2, col = pal["accent3"])
```

::: {.fignote}
*D* is simply the **largest vertical distance** between the two curves. Bigger
*D* = worse agreement. The test converts *D* into a p-value; the null hypothesis
is *"the data come from this distribution"*, so a **small p-value means the fit
is rejected**.
:::

## Running the test

```{r ks-run}
ks <- t(sapply(names(fits), function(nm) {
  e <- fits[[nm]]$estimate
  pf <- c(normal = "pnorm", lognormal = "plnorm", gamma = "pgamma",
          weibull = "pweibull")[nm]
  k <- suppressWarnings(ks.test(yld, pf, e[1], e[2]))
  c(D = unname(k$statistic), p = k$p.value)
}))
knitr::kable(round(ks[order(ks[, "D"]), ], 3))
```

::: {.fignote}
The ranking by *D* agrees with the ranking by AIC here — reassuring, but the two
measure different things: AIC is about the **density at the data points**, *D*
about the **worst mismatch in the cumulative curve**. None of the p-values is
small, so no candidate is formally rejected: with *n* = 60 the test simply has
little power.
:::

## Two things that make that p-value unreliable {.smaller}

::: {.fragment}
**1. Ties.** The test assumes a continuous variable, where no value occurs
twice. Repeated values ("ties") have no exact solution, and R warns you.

```{r ties}
#| echo: true
#| code-fold: false
sum(duplicated(c(immer$Y1, immer$Y2)))
```
:::

::: {.fragment}
**2. Estimated parameters.** We fitted the reference distribution **to the very
data we then test against it**. The curve has been pulled towards the data, so
*D* is too small and the p-value **too optimistic** — the fit looks better than
it is.

The **Lilliefors test** (package `nortest`) is the KS-test corrected for exactly
this. The **Anderson-Darling** test is another alternative, and is less
sensitive to outliers.
:::

## Formal tests for normality

```{r shapiro}
vs <- list(`nlschools$lang` = lang, `nlschools$IQ` = iq,
           `immer yields` = yld, `geyser$waiting` = wait)
st <- t(sapply(vs, function(v) {
  s <- shapiro.test(v); c(n = length(v), W = s$statistic, p = s$p.value)
}))
knitr::kable(signif(st, 3))
```

::: {.fignote}
The **Shapiro-Wilk** test is a dedicated normality test with a good reputation.
*W* near 1 means "close to normal". Note the trap already: `nlschools$lang` looks
perfectly bell-shaped to the eye, yet is rejected with a p-value of 10⁻²¹ — and
`geyser$waiting`, which is blatantly bimodal, is rejected no more decisively.
:::

## Why a normality test is not the answer

```{r shapiro-n}
#| fig-height: 4.0
parset()
set.seed(1)
big <- rlnorm(2000, 0, 0.6)
ns2 <- c(10, 20, 50, 100, 500, 2000)
pv <- sapply(ns2, function(n) shapiro.test(big[1:n])$p.value)
plot(ns2, pv, type = "b", pch = 19, cex = 1.6, log = "x", lwd = 2,
     col = pal["accent"], xlab = "sample size n", ylab = "Shapiro-Wilk p-value",
     ylim = c(0, 1))
abline(h = 0.05, lty = 2, lwd = 2, col = pal["accent2"])
text(12, 0.10, "p = 0.05", col = pal["accent2"], pos = 4, cex = 1.1)
```

::: {.fignote}
The same clearly **non-normal** (log-normal) data throughout. With *n* = 10 the
test cannot tell the difference; by *n* = 500 it is certain. So a
non-significant result on a small sample proves nothing, and on a large sample
*every* real data set is rejected, because no real data set is exactly normal.
The question is never "is it normal?" but "**is it normal enough?**"
:::

## The QQ-plot tells you *how* it deviates

```{r qq}
#| fig-height: 4.2
parset(mfrow = c(1, 4), mar = c(4.0, 4.0, 2.4, 0.8), cex.main = 1.3)
for (nm in names(vs)) {
  qqnorm(vs[[nm]], main = nm, pch = 19, cex = 0.4,
         col = adjustcolor(pal["accent"], 0.5), xlab = "theoretical", ylab = "observed")
  qqline(vs[[nm]], lwd = 2, col = pal["accent2"])
}
```

::: {.fignote}
A p-value gives one number; the QQ-plot gives a diagnosis. Left tail bending
below the line = **left skew** (`lang`); a staircase = **discreteness** (`IQ`);
an S-shape = **heavy tails**; the double bend of `geyser$waiting` betrays
**bimodality**. This is the graphical evaluation Dormann recommends over any
test.
:::

## What you will do in practical 2b {.smaller}

1. build the **likelihood function** for the coin example by hand, find its
   maximum, and see that the log-likelihood peaks in the same place
2. fit a normal distribution to the `Melanoma` survival times; compute the
   likelihood, log-likelihood, AIC and BIC **by hand** and check against R
3. fit normal, log-normal, gamma and Weibull; rank them and draw all four on
   the histogram
4. compare the same fits with the **ECDF** and the **Kolmogorov-Smirnov** test;
   diagnose the ties warning
5. run **Shapiro-Wilk** and QQ-plots on four variables, and demonstrate the
   sample-size trap
6. the tram data: **Poisson or normal?** — compare mean with variance, then BIC
7. `quine$Days`: Poisson versus **negative binomial**, with expected frequencies
8. simulate from a **logistic** distribution, so that for once you know the
   truth, and see whether AIC recovers it

::: {.callout-tip}
Everything you need is `fitdistr()`, `logLik()`, `AIC()`, `BIC()`, `ecdf()`,
`ks.test()`, `shapiro.test()` and `qqnorm()`.
:::

## Take-home messages {.smaller}

1. An **estimator** is a rule, an **estimate** is a number. Good estimators are
   consistent and unbiased.
2. The **likelihood** is the density function read as a function of the
   *parameters* with the data held fixed — the same `d...()` call, the other way
   round.
3. Always use the **log-likelihood**: the product underflows to zero, the sum
   does not, and the logarithm cannot move the maximum.
4. **Maximum likelihood** finds the parameter values at the summit. For the
   normal and Poisson there are formulas (mean, sd); otherwise R searches
   numerically via `fitdistr()`.
5. Report estimates **with their standard errors** — precision grows only with
   the square root of *n*.
6. Compare candidate distributions with $\ell$, but pay for parameters:
   **AIC / AICc / BIC, lower is better**.
7. Check the fit against the **ECDF** as well; the **KS-test** gives *D*, but
   ties and self-estimated parameters make its p-value optimistic (→ Lilliefors).
8. **Never trust a normality test alone.** Small samples pass everything, large
   samples fail everything — look at the QQ-plot.

::: {.callout-tip}
Dormann, C. F. (2020). *Environmental Data Analysis: An Introduction with
Examples in R*, Chapters 3 and 4. Springer.
:::
